---
title: "Decline in Baltic Sea cod body conditon"
author: "Max Lindmark"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format: 
  html:
    embed-resources: true
    fig-width: 8
    toc: true
    #fig-asp: 0.618
knitr: 
  opts_chunk:
    fig.align: center
editor: source
execute: 
  echo: true
  eval: true
  cache: false
---

```{r}
#| warning: false
#| message: false

library(readr)
library(dplyr)
library(rnaturalearth)
library(sf)
library(ggplot2); theme_set(theme_light())
library(viridis)
```

### Introduction
In this vignette, I present showcase the data used in Lindmark *et al.*, (2023) on body condition of Eastern Baltic cod, between years 1993--2021. We use data from DATRAS (<https://datras.ices.dk/>), and the process to clean and prepare data, including adding covariates, can be found here: <https://github.com/maxlindmark/cod-condition/tree/master/R/clean_data>.

Now we'll read the cleaned data. To bypass size limitations on GitHub, I read it in two stages:

```{r}
#| warning: false
#| message: false

d1 <- read_csv("https://raw.githubusercontent.com/maxlindmark/cod-condition/master/data/for_analysis/mdat_cond_(1_2).csv")
d2 <- read_csv("https://raw.githubusercontent.com/maxlindmark/cod-condition/master/data/for_analysis/mdat_cond_(2_2).csv")

d <- bind_rows(d1, d2)
```

We have nearly 100.000 observations of cod condition. In the paper, we develop an spatiotemporally standardize condition index, using geostatistical models.

```{r}
#| warning: false
#| message: false

glimpse(d)
```

### Visualize data

First make a simple map function:

```{r}
#| warning: false
#| message: false

# Specify map ranges
ymin = 52; ymax = 60; xmin = 10; xmax = 24

map_data <- ne_countries(
  scale = "large",
  returnclass = "sf", continent = "europe")

# Crop the polygon for plotting and efficiency:
# st_bbox(map_data) # find the rough coordinates
sf::sf_use_s2(FALSE)
# https://stackoverflow.com/questions/68478179/how-to-resolve-spherical-geometry-failures-when-joining-spatial-data

swe_coast <- suppressWarnings(suppressMessages(
  st_crop(map_data,
          c(xmin = xmin, ymin = ymin, xmax = xmax, ymax = ymax))))

# Transform our map into UTM 33 coordinates, which is the equal-area projection we fit in:
utm_zone33 <- 32633
swe_coast_proj <- sf::st_transform(swe_coast, crs = utm_zone33)

# Make default base map plot
xmin2 <- 303000; xmax2 <- 916000; xrange <- xmax2 - xmin2
ymin2 <- 5980000; ymax2 <- 6450000; yrange <- ymax2 - ymin2

map_plot <- 
ggplot(swe_coast_proj) + 
  xlim(xmin2, xmax2) +
  ylim(ymin2, ymax2) +
  labs(x = "Longitude", y = "Latitude") +
  geom_sf(size = 0.3)
```

Visualize data in space

```{r}
#| warning: false
#| message: false

d_sum <- d |> 
  summarise(n = n(), .by = c(X, Y, year))

map_plot + 
  geom_point(data = d_sum, aes(X*1000, Y*1000, fill = year, size = n),
             alpha = 0.5, shape = 21, stroke = 0.1) +
  scale_fill_viridis(option = "F") +
  scale_size_continuous(
    range = c(0.5, 5)
  )
```

We can also visualize trends over time

```{r}
#| warning: false
#| message: false
#| fig-height: 8

# Calculate a quick le Cren condition factor
a <- exp(coef(lm(log(weight_g) ~ log(length_cm), data = d))[1])
b <- coef(lm(log(weight_g) ~ log(length_cm), data = d))[2]

d <- d |>
  mutate(condition = weight_g / (a*length_cm^b)) |> 
  filter(condition > 0.3 & condition < 2)

d |> 
  summarise(n = n(),
            mean_cond = median(condition),
            .by = c(year, ices_rect)) |> 
  ggplot(aes(year, mean_cond, color = ices_rect, size = n)) + 
  geom_jitter(height = 0, alpha = 0.65) + 
  labs(x = "Year", y = "Mean condition", color = "ICES\nrectangle") +
  theme(legend.position = "bottom") + 
  guides(color = guide_legend(ncol = 8),
         size = guide_legend(ncol = 1)) +
  geom_smooth(aes(year, mean_cond), method = "gam", formula = y~s(x, k=5),
              inherit.aes = FALSE, color = "gray30")
```

To put this into context: a change in condition factor by two units leads to a 16% decline in weight for a cod of 30 cm

```{r}
# Standard weight of a 30 cm cod (W = K*a*L^b)
a*30^b

# For K = 1.2
1.2*a*30^b

(((a*30^b) - (1.2*a*30^b)) / (1.2*a*30^b))*100
```



### References
Lindmark, M., Anderson, S. C., Gogina, M., & Casini, M. (2023). Evaluating drivers of spatiotemporal variability in individual condition of a bottom-associated marine fish, Atlantic cod (Gadus morhua). ICES Journal of Marine Science, 80(5), 1539–1550. <https://doi.org/10.1093/icesjms/fsad084>

### Sessioninfo

```{r}
sessionInfo()
```

